A Comparative Study of Game-Theoretic and Reinforcement Learning Approaches for Jam-Resilient Wireless Networks in IoT

### 2025-05-31 paper/main.tex
Appendix updated

% LaTeX template for "A Comparative Study of Game-Theoretic and Reinforcement Learning Approaches for Jam-Resilient Wireless Networks in IoT"
\documentclass[conference]{IEEEtran}
%======================== Packages ========================
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{algorithm,algpseudocode}
\usepackage{cite}
\usepackage{url}
\usepackage{booktabs} 
\usepackage[utf8]{inputenc} 
\usepackage[T1]{fontenc}  
\usepackage{microtype}

\graphicspath{{figures/}}

@@

\author{
  \IEEEauthorblockN{Muhammad Umar Yaksambi}
  \IEEEauthorblockA{
    Dept. of CSE (Data Sciences)\\
    RV College Of Engineering\\
    Bengaluru, India\\
    Email: umaryaksambi@gmail.com
  }
  \and
  \IEEEauthorblockN{Anuj Devpura}
  \IEEEauthorblockA{
    Dept. of CSE (Data Sciences)\\
    RV College Of Engineering\\
    Bengaluru, India\\
    Email: anujdevpura2004@gmail.com
  }
  \and
  \IEEEauthorblockN{Sadashiv Todakar}
  \IEEEauthorblockA{
    Dept. of CSE (Data Sciences)\\
    RV College Of Engineering\\
    Bengaluru, India\\
    Email: sadashivtodakar@gmail.com
  }
}

@@

Simulations are conducted on Erd\H{o}s–R\'enyi network topologies ($n = 10$, $p = 0.4$) over 300 time steps and five independent trials. Key performance metrics include attacker and defender payoffs, detection rate, and overall network health. Matchup matrices, convergence curves, and statistical validation via ANOVA ($p < 0.001$) are used to assess performance across strategy combinations.

@@

\subsection{Boxplot Comparisons}
Figure~\ref{fig:def_box} shows the distribution of final defender payoffs at $T=300$ across defender strategies (hue indicates attacker type). Bayesian defenders exhibit tighter payoff distributions (IQR = 0.2) compared to Q-Learning (IQR = 0.35), indicating more stable and predictable behavior. Additionally, the median payoff for Bayesian strategies is higher in all attacker pairings, underscoring their robustness against adversarial variability~\cite{zeng2023comparison}.

@@

\subsection{Statistical Analysis}
Table~\ref{tab:anova} summarizes one-way ANOVA results on metrics at the final interval ($\tau=300$). All metrics show $p<0.001$, confirming significant differences between strategy matchups. The high F-statistics for attacker payoff and detection rate indicate that the choice of defense strategy has a substantial influence on these outcomes. The effect size for attacker payoff ($\eta^2 \approx 0.68$) further suggests that the strategy type accounts for a large portion of the observed variance.

@@

%======================== Code Availability ========================
\section*{Code Availability}
The full code, including \texttt{core\_sim.py}, \texttt{run\_experiments.py}, and \texttt{plot\_results.py}, is available at:\\
\url{https://github.com/UmarYaksambi/Adaptive-Defense-Wireless-Networks}. Detailed instructions for replication and dataset generation are provided.

%======================== Appendix ========================
\appendix
\section{Additional Figures}
All supplementary heatmaps (frequency usage distributions and self-play convergence variants) are included in the online repository due to space constraints.

\section{Hyperparameter Grid}
Tested hyperparameter values:
\begin{itemize}
  \item Learning rate $\alpha \in \{0.01,0.05,0.1\}$
  \item Discount factor $\gamma \in \{0.8,0.9,0.99\}$
  \item Exploration $\epsilon_{0}=0.5$, decay $\{0.995,0.998\}$, $\epsilon_{\min}=0.01$
  \item Trials $M=5$, steps $T=300$, logging interval $\tau=50$
\end{itemize}

>>>>

% LaTeX template for "A Comparative Study of Game-Theoretic and Reinforcement Learning Approaches for Jam-Resilient Wireless Networks in IoT"
\documentclass[conference]{IEEEtran}
%======================== Packages ========================
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{algorithm,algpseudocode}
\usepackage{cite}
\usepackage{url}
\usepackage{booktabs} 
\usepackage[utf8]{inputenc} 
\usepackage[T1]{fontenc}  
\usepackage{microtype}
\usepackage[hidelinks]{hyperref}

\graphicspath{{figures/}} % Tells LaTeX to look in the 'figures' subdirectory

@@

\author{
  \IEEEauthorblockN{Muhammad Umar Yaksambi}
  \IEEEauthorblockA{
    Dept. of CSE (Data Sciences)\\
    RV College Of Engineering\\
    Bengaluru, India\\
    Email: mumaryaksambi.cd23@rvce.edu.in
  }
  \and
  \IEEEauthorblockN{Anuj Devpura}
  \IEEEauthorblockA{
    Dept. of CSE (Data Sciences)\\
    RV College Of Engineering\\
    Bengaluru, India\\
    Email: anujdevpura.cd23@rvce.edu.in
  }
  \and
  \IEEEauthorblockN{Sadashiv Todakar}
  \IEEEauthorblockA{
    Dept. of CSE (Data Sciences)\\
    RV College Of Engineering\\
    Bengaluru, India\\
    Email: sadashivtodakar.cd23@rvce.edu.in
  }
}

@@

Simulations are conducted on Erd\H{o}s–R\'enyi network topologies ($n = 10$, $p = 0.4$) over 500 time steps and five independent trials. Key performance metrics include attacker and defender payoffs, detection rate, and overall network health. Matchup matrices, convergence curves, and statistical validation via ANOVA ($p < 0.001$) are used to assess performance across strategy combinations.

@@

\subsection{Boxplot Comparisons}
Figure~\ref{fig:def_box} shows the distribution of final defender payoffs at $T=500$ across defender strategies (hue indicates attacker type). Bayesian defenders exhibit tighter payoff distributions (IQR = 0.2) compared to Q-Learning (IQR = 0.35), indicating more stable and predictable behavior. Additionally, the median payoff for Bayesian strategies is higher in all attacker pairings, underscoring their robustness against adversarial variability~\cite{zeng2023comparison}.

@@

\subsection{Statistical Analysis}
Table~\ref{tab:anova} summarizes one-way ANOVA results on metrics at the final interval ($\tau=500$). All metrics show $p<0.001$, confirming significant differences between strategy matchups. The high F-statistics for attacker payoff and detection rate indicate that the choice of defense strategy has a substantial influence on these outcomes. The effect size for attacker payoff ($\eta^2 \approx 0.68$) further suggests that the strategy type accounts for a large portion of the observed variance.

@@

%======================== Code Availability ========================
\section*{Code Availability}
The full code, including \texttt{core\_sim.py}, \texttt{run\_experiments.py}, and \texttt{plot\_results.py}, is available at our \href{https://github.com/UmarYaksambi/Adaptive-Defense-Wireless-Networks}{GitHub Repository}.

Detailed instructions for replication and dataset generation are provided.

%======================== Appendix ========================
\appendix
\section{Additional Figures}
The following supplementary figures provide further insight into the experimental results and analysis discussed in the main text.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_atk_convergence.pdf}
    \caption{Attacker payoff convergence (all matchups or aggregated).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_atk_convergence_Bayesian_Game_vs_Bayesian_Game.pdf}
    \caption{Attacker payoff convergence: Bayesian Game vs Bayesian Game.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_atk_convergence_Bayesian_Game_vs_Q_Learning.pdf}
    \caption{Attacker payoff convergence: Bayesian Game vs Q-Learning.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_atk_payoff_stdev_boxplot.pdf}
    \caption{Standard deviation of attacker payoff (boxplot).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_collision_count_over_time.pdf}
    \caption{Collision count over time.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_def_convergence.pdf}
    \caption{Defender payoff convergence (all matchups or aggregated).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_def_payoff_stdev_boxplot.pdf}
    \caption{Standard deviation of defender payoff (boxplot).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_detection_events_over_time.pdf}
    \caption{Detection events over time.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_detection_variance_histogram.pdf}
    \caption{Histogram of detection rate variance.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_entropy_atk_Q_learning.pdf}
    \caption{Entropy of attacker strategies (Q-Learning).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_entropy_def_Q_learning.pdf}
    \caption{Entropy of defender strategies (Q-Learning).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_freq_def_QL_vs_Bayes.pdf}
    \caption{Defender strategy/channel usage frequency (Q-Learning vs Bayesian).}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/appendix/fig_health_variance_histogram.pdf}
    \caption{Histogram of network health variance.}
\end{figure}

All supplementary heatmaps (frequency usage distributions and self-play convergence variants) are included in the online repository due to space constraints.

\section{Hyperparameter Grid}
Tested hyperparameter values:
\begin{itemize}
  \item Learning rate $\alpha \in \{0.01,0.05,0.1\}$
  \item Discount factor $\gamma \in \{0.8,0.9,0.99\}$
  \item Exploration $\epsilon_{0}=0.5$, decay $\{0.995,0.998\}$, $\epsilon_{\min}=0.01$
  \item Trials $M=5$, steps $T=500$, logging interval $\tau=50$
\end{itemize}
